Random Forests for Automatic Matching Bullet-Lands

### 2015-12-19 imaging-paper.Rnw
Found and corrected typo
Fix figure reference

\section{Evaluation}
\hh{In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
}
\hh{As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we ues cross sections 25$\mu m$ apart and require a minimum of 0.9 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands shows scratch marks across the surface.}
%
\begin{figure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\linewidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br9 Bullet 2-4}
\includegraphics[width=\linewidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet B-2}
\includegraphics[width=\linewidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet Q-4}
\includegraphics[width=\linewidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}\hh{Images of the four lands that got flagged during the quality assessment. All of them show scratch marks across the striation marks from the barrel. They will be excluded from the remainder of the analysis.}}
\end{figure}
%
\hh{We exclude these four lands from the further matching considerations and 
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are running $15 \times 6 -2 = 90 - 2 = 88$ lands from unkown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). }
 When things go perfectly, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the number of maximum consecutive matching striae between land C-3 and all 120 known lands. Two lands show a high CMS. These correspond to the known matches, shown in Figures~\ref{fig:hamby-perfect}b and c. 
 Unfortunately, not all results are as clear cut. 
It is not reasonable to assume that we can match all lands, but we can try to maximize the number of matches to get an overview of what we can reasonably expect from an automated match.

@@

Another benefit is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}. In a random forest multiple trees are fitted. For each of these trees, only two thirds of the observations are used for fitting, while predictions for the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.7\\linewidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 25 trees, the error rate of land-to-land comparisons stabilizes at 0.0040. This is a weighted average between a false positive error rate of 0.000098 and an error rate of false negatives of 0.22. \hh{This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known matches and non-matches (see Figures~\ref{fig:treeresults} and ~\ref{fig:tree-forest}).}
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched (see Figure~\ref{fig:treeresult}). 
\begin{figure}[hbtp]
<<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.7\\linewidth'>>=
bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
  n = n(),
  pred = sum(pred > 0.5),
  forest = sum(forest > 0.5)
)
library(reshape2)
bm <- melt(data=bullets, id.var=c("bullet", "n"))

>>>>

\section{Evaluation}
\hh{In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
}
\hh{As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we use cross sections 25$\mu m$ apart and require a minimum of 0.9 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands shows scratch marks across the surface.}
%
\begin{figure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\linewidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br9 Bullet 2-4}
\includegraphics[width=\linewidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet B-2}
\includegraphics[width=\linewidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet Q-4}
\includegraphics[width=\linewidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}\hh{Images of the four lands that got flagged during the quality assessment. All of them show scratch marks across the striation marks from the barrel. They will be excluded from the remainder of the analysis.}}
\end{figure}
%
\hh{We exclude these four lands from the further matching considerations and 
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are running $15 \times 6 -2 = 90 - 2 = 88$ lands from unkown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). }
 When things go perfectly, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the number of maximum consecutive matching striae between land C-3 and all 120 known lands. Two lands show a high CMS. These correspond to the known matches, shown in Figures~\ref{fig:hamby-perfect}b and c. 
 Unfortunately, not all results are as clear cut. 
It is not reasonable to assume that we can match all lands, but we can try to maximize the number of matches to get an overview of what we can reasonably expect from an automated match.

@@

Another benefit is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}. In a random forest multiple trees are fitted. For each of these trees, only two thirds of the observations are used for fitting, while predictions for the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.7\\linewidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 25 trees, the error rate of land-to-land comparisons stabilizes at 0.0040. This is a weighted average between a false positive error rate of 0.000098 and an error rate of false negatives of 0.22. \hh{This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known matches and non-matches (see Figures~\ref{fig:treeresult} and ~\ref{fig:tree-forest}).}
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched (see Figure~\ref{fig:treeresult}). 
\begin{figure}[hbtp]
<<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.7\\linewidth'>>=
bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
  n = n(),
  pred = sum(pred > 0.5),
  forest = sum(forest > 0.5)
)
library(reshape2)
bm <- melt(data=bullets, id.var=c("bullet", "n"))
